Practical Optional Types for Clojure

### 2015-02-27 formal-model.tex
update figure
move syntax descriptions to overview

%TODO EXPLAIN what is a proposition?
The first insight about occurrence typing is that
logical formulas
can be used to represent type information about our programs
by relating parts of the runtime environment to types
via propositional logic.
\emph{Type Propositions} \prop{} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
``variable \clj{x} is not \clj{nil}''---in our logical system we write these as
{\isprop{\NumberFull}{\x{}}}
and {\notprop{\Nil{}}{\x{}}}. 
The other propositions are standard logical connectives: implications, conjunctions,
disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions
(figure~\ref{main:figure:termsyntax}).

@@

Given a set of propositions, we can use logical reasoning to derive
new information about our programs
with the judgement \inpropenv{\propenv{}}{\prop{}}.
In addition to the standard rules for the logical connectives, the key
rule is L-Update, which combines multiple propositions about the same variable,
allowing us to refine its type.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover L-Update in more detail in section~\ref{sec:formalpaths}.

Finally, this approach allows the type system to track
programming idioms from 
dynamic languages
using implicit type-based reasoning based on the result of
conditional tests.
For instance, example~\ref{example:conditionalflow}
only utilises \clj{x} once
the programmer is convinced it is safe to do so based whether
\clj{(number? x)}
is 
true or false. 
To express this in the type system, every expression 
is described by two propositions: a `then' proposition
for when it reduces to a true value, and an `else' proposition
when it reduces to a false value---for \clj{(number? x)}
the then proposition is {\isprop{\NumberFull}{\x{}}} and 
the else proposition is {\notprop{\NumberFull}{\x{}}}.
%\ref{main:figure:typingrules}

@@

Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals and let expressions.
All binding forms introduce fresh variables.
Values include booleans, \nil{}, class literals, keywords, 
numbers,
constants and closures. 
Value environments map local bindings to values.

@@

Figure~\ref{main:figure:typingrules} contains the core typing rules.
The key rule for reasoning about conditional control flow is
T-If.

@@

The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
are used as assumptions in the then and else branch respectively.
If the result of the \ifliteral{} is true, then the value either
came from \e{2}, in which case \thenprop{\prop{2}} is true, or from \e{3},
which implies \thenprop{\prop{3}} is true. 
The else proposition is \orprop{\elseprop{\prop{2}}}{\elseprop{\prop{3}}} 
similarly.
The T-Local rule connects the type system to the proof system over type propositions
to derive a type for a variable.
Using this rule, the type system can then appeal to L-Update to refine the type
assigned to \x{}.

We are now equipped to type check 
example~\ref{example:conditionalflow}.
The test expression uses T-App, 
$$
\judgement{\isprop{\UnionNilNum{}}{\x{}}}{\hastype{\appexp{\numberhuh{}}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
since \numberhuh{} has type
{\ArrowOne{\x{}}{\Top}{\Boolean}
        {\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}}
      and \x{} has object \x{}.

@@

Subtyping (figure~\ref{main:figure:subtyping}) 
is a reflexive and transitive relation with top type \Top. 
Singleton types are instances of their respective classes---boolean singleton types
are of type \Boolean{}, class literals are instances of \Class{} and keywords are
instances of \Keyword{}.
Instances of classes \class{} are subtypes of \Object{}. Since function types 
are subtypes of \IFn{}, all types except for \Nil{} are subtypes of \Object{},
so \Top{} = {\Union{\Nil}{\Object}}.
Function subtyping is contravariant left of the arrow---latent propositions, object
and the result type are covariant.
Subtyping for untagged unions is standard.

@@

Recall example~\ref{example:doexception}.
\begin{minted}{clojure}
...  (do (if (number? x) nil (throw (Exception.)))
         (inc x)) ...
\end{minted}

@@

We now present \HMapliteral{} types in detail.
Figure~\ref{main:figure:hmapsyntax} extends the syntax for terms and types.
The syntax 
$$
\clj{(HMap :mandatory p :absent-keys a :complete? b)}
$$
we introduced in section~\ref{sec:overview}
is now written \HMapgeneric{\mandatory{}}{\absent{}},
where \completenessmeta{} is {\complete{}} if \clj{b} is \clj{true} and is {\partial{}} otherwise. 
The \mandatory{} entry specifies a set of known present keyword entries and \absent{} tracks known absent keys.
To ease presentation, 
if a HMap is declared as \complete{} then \absent{} implicitly contains all other keywords not specified in \mandatory{}.
%\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. 
Keys cannot be both present and absent.

@@

Subtyping for HMaps (figure~\ref{main:figure:hmapsyntax})
designate \MapLiteral{} as a common supertype for all HMaps.
S-HMap says that a HMap is a subtype of another HMap if they agree
on \completenessmeta{}, agree on mandatory entries with subtyping
and at least cover the absent keys of the supertype.
Complete maps are subtypes of partial maps
as long as they agree on the mandatory entries of the partial map via subtyping (S-HMapP).

Figure~\ref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
We only handle unions of HMaps in T-GetHMap, where every element in the union
contains the key we are looking up, in particular to support operations like
looking up \clj{:op} on a map of type \clj{Expr} (example~\ref{example:decleaf}).
The object in each T-Get rule is now more complicated than in T-Local---the 
next section discusses this in detail.
T-Assoc extends a HMap with a mandatory entry.

Figure~\ref{main:figure:hmapsyntax} contains the semantics for \getliteral{}
and \assocliteral{}.
The B-Get rule evaluates first evaluates the map and the key, asserts
the key must be present in the map and returns the associated key. If
the entry is missing, B-GetMissing says we return \nil{}.

@@

\paragraph{Key path element} We introduce our first path element
{\keype{\k{}}}, which represents the operation of looking up
a key \k{}.
We directly relate this to our typing rule T-GetHMap
(figure~\ref{main:figure:hmapsyntax}) by
checking the then branch of the first conditional test is checked in 
an equivalent version of example~\ref{example:decleaf}.
\begin{minted}{clojure}
  (fn [m :- Expr]
    (if (= (get m :op) :if)
      {:op :if, ...}
      (if ...)))
\end{minted}

We do not specifically support \equivliteral{} in our calculus, 
but on keyword arguments it works identically to \clj{isa?} which we model
in section~\ref{sec:isaformal}.
Intuitively, if {\judgement{\propenv{}}{\hastype{\e{}}{\t{}}}{\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}{\object{}}}
then \equivapp{\e{}}{\makekw{if}} has the proposition set 
$$
{\replacefor{\filterset{\isprop{\Value{\makekw{if}}}{\x{}}}{\notprop{\Value{\makekw{if}}}{\x{}}}}{\object{}}{\x{}}}
$$
where substitution reduces to \topprop{} if \object{} = \emptyobject{}.

@@

\paragraph{Class path element} Our second path element \classpe{} is used in the latent
object of the constant \classconst{} function. Like Clojure's \clj{class}
function \classconst{} returns the argument's class or \nil{}
if passed \nil{}.
$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$
\begin{mathpar}
\constanttypefigure{}
\end{mathpar}
The full semantics are given in figure~\ref{main:figure:primitivesem}.
The definition of \updateliteral{} supports various idioms relating to \classpe{}
which we introduce in section~\ref{sec:isaformal}.

@@

In section~\ref{sec:overviewjavainterop} we discussed the role of type hints
to help eliminate reflective calls.
In this section, we introduce our model of Java and provide user-facing
syntax corresponding to Clojure's Java interoperability forms and type hinted forms.
Then we model the Clojure compiler's compile-time reflection resolution
algorithm.
To achieve this, first we define notation for
\emph{non-reflective} Java forms that 
unambiguously call a field, method or constructor.
Then we define a rewrite relation that uses
type hints to resolve reflection explicitly.
Finally we give typing rules that model how
Typed Clojure interacts with non-reflective calls.

@@

We extend the syntax with type hinted expressions, Java field lookups and calls to
methods and constructors. We model the syntax after the `dot' special
form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
The Java expressions come without typing rules because Typed Clojure
only reasons about resolved reflection
(as demonstrated in section~\ref{sec:overviewjavainterop}).

Now we model the compiler's reflection-resolution algorithm.
Figure~\ref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
The method ${\methodstaticexp {\classhint{1}}
                             {\overrightarrow {\classhint{i}}}
                             {\classhint{2}}
                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
is a non-reflective call to the \mth{} method on class {\class{1}}, 
with Java signature 
${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
calls the field on class {\classhint{1}} with Java signature
${\classhint{2}}\ \fld{};$.
The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                               {\class{}} {\overrightarrow{\e{i}}}}
calls the constructor with Java signature
${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.

@@

For demonstration purposes let us rewrite
a simple field expression, with the assumptions that
that we have a class \Point{} with a field \getx{} of Java type \Number
and \intaenv {\taenv{}} {\x{}} {\Point{}}.
In rewriting
the expression \fieldexp{\getx}{\x{}},
{\Point{}} is inferred as the type hint of \x{}
and \Number{} is the field type
by \fieldtypeliteral{} (figure~\ref{main:figure:javatyping}).
Now we just plug in the new information into our new expression
$$
\fieldstaticexp{\Point}{\Number}{\getx}{\x{}}.
$$

@@

The other rules T-MethodStatic and T-NewStatic work similarly (figure~\ref{main:figure:javatyping}), varying
in the choice of nilability in the conversion function---method returns
are nilable and constructor returns are non-nilable.

The operational semantics B-Field, B-New and B-Method (figure~\ref{main:figure:javatyping}) simply evaluate their
arguments and call the relevant JVM operation---section~\label{sec:metatheory}
states our exact assumptions about each.

@@

We now consider the core dispatch mechanism for multimethods. 
Recalling the examples in section~\ref{sec:multioverview},
\isaliteral{} is
a subclassing test for classes, otherwise an equality test---we do not model
the semantics for vectors.

The main interest for the typing rule is \isacompareliteral{} 
(figure~\ref{main:figure:mmsyntax}).

@@

To demonstrate,
\mintinline{clojure}{(isa? (class x) Keyword)}
has the proposition set \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}}.

@@

To ease presentation, we present \emph{immutable}
multimethods (figure~\ref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. Example~\ref{example:rep} is now written
\begin{minted}{clojure}
(let [rep (defmulti [Any -> String] class)
      rep (defmethod rep Keyword [x] (str (name x)))
      rep (defmethod rep Number [x] (str (inc x)))]
  (rep :a)) ;=> "a"
\end{minted}

Multimethod semantics are in figure~\ref{main:figure:mmsyntax}.
B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
B-DefMethod returns a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
which \getmethodliteral{} uses to choose the method to invoke.

The typing rules (figure~\ref{main:figure:mmsyntax}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
where \s{} is the interface type for beta reduction, and \t{} is the type for
the dispatch function. T-DefMulti uses the provided interface type and dispatch function
to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
that must be true if this method is being invoked. The example can be type checked without
changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
method bodies. T-App handles beta reduction because multimethods are subtypes of functions
(figure~\ref{main:figure:mmsyntax}).

>>>>

%TODO EXPLAIN what is a proposition?
The first insight about occurrence typing is that
logical formulas
can be used to represent type information about our programs
by relating parts of the runtime environment to types
via propositional logic.
\emph{Type Propositions} \prop{} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
``variable \clj{x} is not \clj{nil}''---in our logical system we write these as
{\isprop{\NumberFull}{\x{}}}
and {\notprop{\Nil{}}{\x{}}}. 
The other propositions are standard logical connectives: implications, conjunctions,
disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions
(\figref{main:figure:termsyntax}).

@@

Given a set of propositions, we can use logical reasoning to derive
new information about our programs
with the judgement \inpropenv{\propenv{}}{\prop{}}.
In addition to the standard rules for the logical connectives, the key
rule is L-Update, which combines multiple propositions about the same variable,
allowing us to refine its type.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover L-Update in more detail in \secref{sec:formalpaths}.

Finally, this approach allows the type system to track
programming idioms from 
dynamic languages
using implicit type-based reasoning based on the result of
conditional tests.
For instance, \egref{example:conditionalflow}
only utilises \clj{x} once
the programmer is convinced it is safe to do so based whether
\clj{(number? x)}
is 
true or false. 
To express this in the type system, every expression 
is described by two propositions: a `then' proposition
for when it reduces to a true value, and an `else' proposition
when it reduces to a false value---for \clj{(number? x)}
the then proposition is {\isprop{\NumberFull}{\x{}}} and 
the else proposition is {\notprop{\NumberFull}{\x{}}}.
%\ref{main:figure:typingrules}

@@

Syntax is given in \figref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals and let expressions.
All binding forms introduce fresh variables.
Values include booleans, \nil{}, class literals, keywords, 
numbers,
constants and closures. 
Value environments map local bindings to values.

@@

\figref{main:figure:typingrules} contains the core typing rules.
The key rule for reasoning about conditional control flow is
T-If.

@@

The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
are used as assumptions in the then and else branch respectively.
If the result of the \ifliteral{} is true, then the value either
came from \e{2}, in which case \thenprop{\prop{2}} is true, or from \e{3},
which implies \thenprop{\prop{3}} is true. 
The else proposition is \orprop{\elseprop{\prop{2}}}{\elseprop{\prop{3}}} 
similarly.
The T-Local rule connects the type system to the proof system over type propositions
via \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}
to derive a type for a variable.
Using this rule, the type system can then appeal to L-Update to refine the type
assigned to \x{}.

We are now equipped to type check 
\egref{example:conditionalflow}:
$$
\clj{(if (number? x) (inc x) 0)}
$$

The test expression uses T-App, 
$$
\judgement{\isprop{\UnionNilNum{}}{\x{}}}{\hastype{\appexp{\numberhuh{}}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
since \numberhuh{} has type
{\ArrowOne{\x{}}{\Top}{\Boolean}
        {\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}}
      and \x{} has object \x{}.

@@

Subtyping (\figref{main:figure:subtyping}) 
is a reflexive and transitive relation with top type \Top. 
Singleton types are instances of their respective classes---boolean singleton types
are of type \Boolean{}, class literals are instances of \Class{} and keywords are
instances of \Keyword{}.
Instances of classes \class{} are subtypes of \Object{}. Since function types 
are subtypes of \IFn{}, all types except for \Nil{} are subtypes of \Object{},
so \Top{} = {\Union{\Nil}{\Object}}.
Function subtyping is contravariant left of the arrow---latent propositions, object
and the result type are covariant.
Subtyping for untagged unions is standard.

@@

Recall \egref{example:doexception}.
\begin{minted}{clojure}
...  (do (if (number? x) nil (throw (Exception.)))
         (inc x)) ...
\end{minted}

@@

We now present \HMapliteral{} types in detail.
\figref{main:figure:hmapsyntax} extends the syntax for terms and types.
The syntax 
$$
\clj{(HMap :mandatory p :absent-keys a :complete? b)}
$$
we introduced in \secref{sec:overview}
is now written \HMapgeneric{\mandatory{}}{\absent{}},
where \completenessmeta{} is {\complete{}} if \clj{b} is \clj{true} and is {\partial{}} otherwise. 
The \mandatory{} entry specifies a set of known present keyword entries and \absent{} tracks known absent keys.
To ease presentation, 
if a HMap is declared as \complete{} then \absent{} implicitly contains all other keywords not specified in \mandatory{}.
%\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. 
Keys cannot be both present and absent.

@@

Subtyping for HMaps (\figref{main:figure:hmapsyntax})
designate \MapLiteral{} as a common supertype for all HMaps.
S-HMap says that a HMap is a subtype of another HMap if they agree
on \completenessmeta{}, agree on mandatory entries with subtyping
and at least cover the absent keys of the supertype.
Complete maps are subtypes of partial maps
as long as they agree on the mandatory entries of the partial map via subtyping (S-HMapP).

\figref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
We only handle unions of HMaps in T-GetHMap, where every element in the union
contains the key we are looking up, in particular to support operations like
looking up \clj{:op} on a map of type \clj{Expr} (\egref{example:decleaf}).
The object in each T-Get rule is now more complicated than in T-Local---the 
next section discusses this in detail.
T-Assoc extends a HMap with a mandatory entry.

\figref{main:figure:hmapsyntax} contains the semantics for \getliteral{}
and \assocliteral{}.
The B-Get rule evaluates first evaluates the map and the key, asserts
the key must be present in the map and returns the associated key. If
the entry is missing, B-GetMissing says we return \nil{}.

@@

\paragraph{Key path element} We introduce our first path element
{\keype{\k{}}}, which represents the operation of looking up
a key \k{}.
We directly relate this to our typing rule T-GetHMap
(\figref{main:figure:hmapsyntax}) by
checking the then branch of the first conditional test is checked in 
an equivalent version of \egref{example:decleaf}.
\begin{minted}{clojure}
  (fn [m :- Expr]
    (if (= (get m :op) :if)
      {:op :if, ...}
      (if ...)))
\end{minted}

We do not specifically support \equivliteral{} in our calculus, 
but on keyword arguments it works identically to \clj{isa?} which we model
in \secref{sec:isaformal}.
Intuitively, if {\judgement{\propenv{}}{\hastype{\e{}}{\t{}}}{\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}{\object{}}}
then \equivapp{\e{}}{\makekw{if}} has the proposition set 
$$
{\replacefor{\filterset{\isprop{\Value{\makekw{if}}}{\x{}}}{\notprop{\Value{\makekw{if}}}{\x{}}}}{\object{}}{\x{}}}
$$
where substitution reduces to \topprop{} if \object{} = \emptyobject{}.

@@

\paragraph{Class path element} Our second path element \classpe{} is used in the latent
object of the constant \classconst{} function. Like Clojure's \clj{class}
function \classconst{} returns the argument's class or \nil{}
if passed \nil{}.
$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$
\begin{mathpar}
\constanttypefigure{}
\end{mathpar}
The full semantics are given in \figref{main:figure:primitivesem}.
The definition of \updateliteral{} supports various idioms relating to \classpe{}
which we introduce in \secref{sec:isaformal}.

@@

In \secref{sec:overviewjavainterop} we discussed the role of type hints
to help eliminate reflective calls.
In this section, we introduce our model of Java and provide user-facing
syntax corresponding to Clojure's Java interoperability forms and type hinted forms.
Then we model the Clojure compiler's compile-time reflection resolution
algorithm.
To achieve this, first we define notation for
\emph{non-reflective} Java forms that 
unambiguously call a field, method or constructor.
Then we define a rewrite relation that uses
type hints to resolve reflection explicitly.
Finally we give typing rules that model how
Typed Clojure interacts with non-reflective calls.

@@

We extend the syntax with type hinted expressions, Java field lookups and calls to
methods and constructors. We model the syntax after the `dot' special
form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
The Java expressions come without typing rules because Typed Clojure
only reasons about resolved reflection
(as demonstrated in \secref{sec:overviewjavainterop}).

Now we model the compiler's reflection-resolution algorithm.
\figref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
The method ${\methodstaticexp {\classhint{1}}
                             {\overrightarrow {\classhint{i}}}
                             {\classhint{2}}
                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
is a non-reflective call to the \mth{} method on class {\class{1}}, 
with Java signature 
${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
calls the field on class {\classhint{1}} with Java signature
${\classhint{2}}\ \fld{};$.
The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                               {\class{}} {\overrightarrow{\e{i}}}}
calls the constructor with Java signature
${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.

@@

For demonstration purposes let us rewrite
a simple field expression, with the assumptions that
that we have a class \Point{} with a field \getx{} of Java type \Number
and \intaenv {\taenv{}} {\x{}} {\Point{}}.
In rewriting
the expression \fieldexp{\getx}{\x{}},
{\Point{}} is inferred as the type hint of \x{}
and \Number{} is the field type
by \fieldtypeliteral{} (\figref{main:figure:javatyping}).
Now we just plug in the new information into our new expression
$$
\fieldstaticexp{\Point}{\Number}{\getx}{\x{}}.
$$

@@

The other rules T-MethodStatic and T-NewStatic work similarly (\figref{main:figure:javatyping}), varying
in the choice of nilability in the conversion function---method returns
are nilable and constructor returns are non-nilable.

The operational semantics B-Field, B-New and B-Method (\figref{main:figure:javatyping}) simply evaluate their
arguments and call the relevant JVM operation---\secref{sec:metatheory}
states our exact assumptions about each.

@@

We now consider the core dispatch mechanism for multimethods. 
Recalling the examples in \secref{sec:multioverview},
\isaliteral{} is
a subclassing test for classes, otherwise an equality test---we do not model
the semantics for vectors.

The main interest for the typing rule is \isacompareliteral{} 
(\figref{main:figure:mmsyntax}).

@@

To demonstrate,
\isaapp{\appexp{\classconst{}}{\x{}}}{\Keyword}
has the proposition set \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}}.

@@

To ease presentation, we present \emph{immutable}
multimethods (\figref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. \egref{example:rep} is now written
\begin{minted}{clojure}
(let [rep (defmulti [Any -> String] class)
      rep (defmethod rep Keyword [x] (str (name x)))
      rep (defmethod rep Number [x] (str (inc x)))]
  (rep :a)) ;=> "a"
\end{minted}

Multimethod semantics are in \figref{main:figure:mmsyntax}.
B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
B-DefMethod returns a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
which \getmethodliteral{} uses to choose the method to invoke.

The typing rules (\figref{main:figure:mmsyntax}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
where \s{} is the interface type for beta reduction, and \t{} is the type for
the dispatch function. T-DefMulti uses the provided interface type and dispatch function
to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
that must be true if this method is being invoked. The example can be type checked without
changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
method bodies. T-App handles beta reduction because multimethods are subtypes of functions
(\figref{main:figure:mmsyntax}).

### 2015-02-27 intro.tex
move syntax descriptions to overview

The popularity of dynamically-typed languages languages in software
development, combined with a recognition that types often improve
programmer productivity, software reliability, and performance, has
led to the recent development of a wide variety of optional and
gradual type systems aimed at checking existing programs written in
existing languages.  These include Microsoft's TypeScript for
JavaScript, Facebook's Hack for PHP and Flow for JavaScript, and MyPy
for Python among the optional systems, and Typed Racket, Reticulated
Python, and GradualTalk among gradually-typed systems.\footnote{We
  reserve the term ``gradual typing'' for systems such as Typed Racket which soundly
  interoperate between typed and untyped code; systems like Typed Clojure or TypeScript which do not
  enforce type invariants we describe as ``optionally typed''.}
%FIXME: add citations for all those systems.

@@

The essence of Typed Clojure, of course, is Clojure, a dynamically
typed language in the Lisp family built to run on the Java Virtual
Machine (JVM) which has recently gained popularity as an alternative
JVM language.  It offers the flexibility of a Lisp dialect, including
macros, emphasizes a functional style via a
standard library of immutable data structures, and provides
interoperability with existing Java code, allowing programmers to use
existing Java libraries without leaving Clojure.
%
Since its initial release in 2007, Clojure has been widely adopted for
``backend'' development in places where its support for paralleism,
functional programming, and Lisp-influenced abstraction is desired on
the JVM. As a result, it now has an extensive base of existing untyped
programs, whose developers can now benefit from Typed Clojure. As
a result, Typed Clojure is used in industry, experience we discuss
in this paper.

@@

In the remainder of this paper, we describe how Typed Clojure's
central innovations, including Java interoperability, multimethods,
and heterogeneously-typed immutable maps, enable this example and many
others. We begin with an example-driven presentation of the main type
system features in Section~\ref{sec:overview}. We then incrementatlly
present a core calculus for Typed Clojure covering all of these
features together in Section~\ref{sec:formal} and prove type
soundness(section \ref{sec:sound}). We then discuss the full
implementation of Typed Clojure, dubbed \coretyped{}, which extends
the formal model in many ways, and the experience gained from its use
in Section~\ref{sec:exp}. Finally, we discuss related work and
conclude.

>>>>

The popularity of dynamically-typed languages in software
development, combined with a recognition that types often improve
programmer productivity, software reliability, and performance, has
led to the recent development of a wide variety of optional and
gradual type systems aimed at checking existing programs written in
existing languages.  These include Microsoft's TypeScript for
JavaScript, Facebook's Hack for PHP and Flow for JavaScript, and MyPy
for Python among the optional systems, and Typed Racket, Reticulated
Python, and GradualTalk among gradually-typed systems.\footnote{We
  reserve the term ``gradual typing'' for systems such as Typed Racket which soundly
  interoperate between typed and untyped code; systems like Typed Clojure or TypeScript which do not
  enforce type invariants we describe as ``optionally typed''.}
%FIXME: add citations for all those systems.

@@

The essence of Typed Clojure, of course, is Clojure, a dynamically
typed language in the Lisp family built to run on the Java Virtual
Machine (JVM) which has recently gained popularity as an alternative
JVM language.  It offers the flexibility of a Lisp dialect, including
macros, emphasizes a functional style via a
standard library of immutable data structures, and provides
interoperability with existing Java code, allowing programmers to use
existing Java libraries without leaving Clojure.
%
Since its initial release in 2007, Clojure has been widely adopted for
``backend'' development in places where its support for parallelism,
functional programming, and Lisp-influenced abstraction is desired on
the JVM. As a result, it now has an extensive base of existing untyped
programs, whose developers can now benefit from Typed Clojure. As
a result, Typed Clojure is used in industry, experience we discuss
in this paper.

@@

In the remainder of this paper, we describe how Typed Clojure's
central innovations, including Java interoperability, multimethods,
and heterogeneously-typed immutable maps, enable this example and many
others. We begin with an example-driven presentation of the main type
system features in \secref{sec:overview}. We then incrementally
present a core calculus for Typed Clojure covering all of these
features together in \secref{sec:formal} and prove type
soundness (\secref{sec:metatheory}). We then discuss the full
implementation of Typed Clojure, dubbed \coretyped{}, which extends
the formal model in many ways, and the experience gained from its use
in \secref{sec:experience}. Finally, we discuss related work and
conclude.
